---
#layout: preprint
layout: review, 11pt
linenumbers: true
title: "rgbif: a package for working with species occurrence data in R"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
address:
  - code: cstar
    address: |
      University of California, Berkeley, CA, USA
abstract: |
      1. xxx

      2. xxx

      3. xxx

      4. xxxx

bibliography: components/references.bib
csl: components/methods-in-ecology-and-evolution.csl
documentclass: components/elsarticle

output:
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: true
    fig_caption: true
---


```{r compile-settings, include=FALSE}
library("methods")
library("knitr")
opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE,
               cache = 1, comment = NA, verbose = TRUE)
basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile'))
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
```

# Introduction

Users of the popular statistical and mathematical computing platform
R [@R] enjoy a wealth of readily installable comparative phylogenetic
methods and tools.

To make the capabilities of NeXML available to R users in an easy-to-use
form, and to lower the hurdles to adoption of the standard, we present
RNeXML, an R package that aims to provide easy programmatic access to
reading and writing NeXML documents, tailored for the kinds of use-cases
that will be common for users and developers of the wealth of evolutionary
analysis methods within the R ecosystem.

```{r echo=FALSE, results='hide', eval=FALSE}
install.packages("rgbif", dependencies=TRUE)
```

```{r echo=FALSE, results='hide'}
library("rgbif")
```

# The rgbif package

The `rgbif` package ...

# Conclusions and future directions

`rgbif` ...

## Acknowledgements

This project was supported in part by the Alfred P Sloan
Foundation (Grant 2013-6-22).

## Data Accessibility

All software, scripts and data used in this paper can be found in the permanent
data archive Zenodo under the digital object identifier (DOI). This DOI corresponds to a snapshot of the GitHub repository at [github.com/sckott/msrgbif](https://github.com/sckott/msrgbif).

# References
