---
#layout: preprint
layout: review, 11pt
linenumbers: true
title: "rgbif: a package for working with species occurrence data in R"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
address:
  - code: cstar
    address: |
      University of California, Berkeley, CA, USA
abstract: |
      1. xxx

      2. xxx

      3. xxx

      4. xxxx

bibliography: components/references.bib
csl: components/methods-in-ecology-and-evolution.csl
documentclass: components/elsarticle

output:
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: true
    fig_caption: true
---


```{r compile-settings, include=FALSE}
library("methods")
library("knitr")
opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE,
               cache = 1, comment = NA, verbose = TRUE)
basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile'))
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
```

# Introduction

Biodiversity loss is one of the greatest challenges of our
time [@Pimm_2014]. Some have called this the sixth great mass
extinction [@Ceballos_2015]. Given this challenge there is a
great need for, and much research is done on, biodiversity
science. The pool of questions about biodiversity we can answer
effectively becomes much larger with more and better data.

The Global Biodiversity Information Facility (hereafter, GBIF,
http://www.gbif.org/) is the largest collection of biodiversity
records globally, currently with 580 million records, 1.6 million
taxa, 15,000 datasets from 770 publishers. In addition, large
biodiversity warehouses such as iNaturalist, VertNet, and USGS's
BISON all feed into GBIF.

Herein, we describe a library (rgbif [@rgbif]) for working with
GBIF data in the R programming environment [@R]. R is an extremely
widely used language in academia, non-profits, and the private sector.
Importantly, R make it easy to do all of the steps of the
research process, including data management, data manipulation
and cleaning, statistics, and vizualization. An R client
for getting GBIF is a powerful tool for reproducible research.

# Install rgbif

```{r echo=FALSE, results='hide', eval=FALSE}
install.packages("rgbif", dependencies=TRUE)
```

```{r echo=FALSE, results='hide'}
library("rgbif")
```

# The rgbif package

The `rgbif` package ...

## Registry

The GBIF reqistry API services are spread across four major sets
of functions:

* Datasets
* Installations
* Networks
* Nodes
* Organizations

### Datasets

Search for datasets

```{r}
dataset_search(keyword="france", return = "data")
```

Get dataset metrics

```{r}
res <- dataset_metrics(uuid='66dd0960-2d7d-46ee-a491-87b9adcfe7b1')
data.frame(names(res$countByRank), unname(unlist(res$countByRank)))
```

### Networks, nodes, and installations

Here, we search for the first give GBIF networks, returning just the
key and title fields.

```{r}
networks(limit=5)$data[ , c('key', 'title')]
```

## Species

## Occurrences

GBIF provides two ways to get occurrence data: through the
/occurrence/search route (see `occ_search`), or via the
/occurrence/download route (many functions, see below).
`occ_search` is more appropriate for smaller data, while
occ_download*() functions are more appropriate for larger
data requests.

### Download API

The download API syntax is similar to the occurrence search API
in that the same parameters are used, but the syntas is quite different.
For example, in the download API you can do greater than searches,
whereas you can not do that in the occurrence search API.

```{r}
occ_download_get("0000066-140928181241064", overwrite = TRUE)
```

### Search API

# Conclusions and future directions

## Acknowledgements

This project was supported in part by the Alfred P Sloan
Foundation (Grant 2013-6-22).

## Data Accessibility

All software, scripts and data used in this paper can be found in the permanent
data archive Zenodo under the digital object identifier (DOI). This DOI corresponds to a snapshot of the GitHub repository at [github.com/sckott/msrgbif](https://github.com/sckott/msrgbif).

# References
