```{r include=FALSE}
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  comment = '#>',
  collapse = TRUE,
  fig.width = 8,
  fig.height = 6,
  cache.path = "cache/"
)
```

In this example, we collect specimen records across different cities using GBIF data from the `rgbif` package.

_Load libraries_

```{r load}
library("rgbif")
library("ggplot2")
library("plyr")
library("httr")
library("RColorBrewer")
library("wicket")
```

_Get bounding boxes for some cites_

Bounding lat/long data is from <https://raw.github.com/amyxzhang/boundingbox-cities/master/boundbox.txt>.

```{r box}
url <- 'https://raw.githubusercontent.com/amyxzhang/
boundingbox-cities/master/boundbox.txt'
rawdat <- content(GET(sub("\n", "", url)), as = "text")
dat <- read.table(
  text = rawdat, header = FALSE, 
  sep = "\t", col.names = c("city","minlat","maxlon","maxlat","minlon"),
  stringsAsFactors = FALSE)
dat <- data.frame(
  city = dat$city, minlon = dat$minlon, 
  minlat = dat$minlat, maxlon = dat$maxlon, 
  maxlat = dat$maxlat,
  stringsAsFactors = FALSE
)
```

A helper function to get count data. GBIF has a count API, but we can't use
that with a geometry search as that API doesn't support geospatial search.
We can however use the search API via `occ_search()` and set `limit = 1` so 
that we 

```{r newsearch}
getdata <- function(x){
  coords <- as.numeric(x[c('minlon','minlat','maxlon','maxlat')])
  wkt <- wicket::wkt_correct(wicket::bounding_wkt(values = coords))
  num <- occ_search(geometry = wkt, limit = 1)$meta$count
  data.frame(
    city = x['city'], 
    richness = num, 
    stringsAsFactors = FALSE
  )
}
```

```{r cache=TRUE}
out <- apply(dat, 1, getdata)
```

_Merge to original table_

```{r}
out <- merge(dat, ldply(out), by = "city")
```

_Add centroids from bounding boxes_

```{r}
out <- transform(out, lat = (minlat + maxlat)/2, lon = (minlon + maxlon)/2)
```

_Plot data_

```{r}
mapp <- map_data('world')
ggplot(mapp, aes(long, lat)) +
  geom_polygon(aes(group=group), fill="white", alpha=0, color="black", size=0.4) +
  geom_point(data=out, aes(lon, lat, color=richness), size=5, alpha=0.8) +
  scale_color_continuous(low = "#60E1EE", high = "#0404C8") +
  labs(x="", y="") +
  theme_grey(base_size=14) +
  theme(legend.position = "bottom", legend.key = element_blank()) +
  guides(color = guide_legend(keywidth = 2))
```
